Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes

### 2017-07-11 pasp/paperclimate.tex
ref version

\begin{abstract}
We set out to evaluate the potential of the Colombian Andes for millimeter-wave astronomical observations. Previous studies for astronomical site testing in this region have suggested that nighttime humidity and cloud cover conditions make most sites unsuitable for professional visible-light observations. Millimeter observations can be done during the day, but require that the precipitable water vapor column above a site stays below $\sim$10 mm. Due to a lack of direct radiometric or radiosonde measurements, we present a method for correlating climate data from weather stations to sites with a low precipitable water vapor column. We use unsupervised learning techniques to low-dimensionally embed climate data (precipitation, rain days, relative humidity, and sunshine duration) in order to group together stations with similar long-term climate behavior. The data were taken over a period of 30 years by 2046 weather stations across the Colombian territory. We find 6 regions with unusually dry, clear-sky conditions, ranging in elevations from 2200 to 3800 masl. We evaluate the suitability of each region using a quality index derived from a Bayesian probabilistic analysis of the station type and elevation distributions. Two of these regions show a high probability of having an exceptionally low precipitable water vapor column. We compared our results with global precipitable water vapor maps and find a plausible geographical correlation with regions with low water vapor columns ($\sim15$ mm) at an accuracy of $\sim20$ km. Our methods can be applied to similar datasets taken in other countries as a first step toward astronomical site evaluation.
\end{abstract}

@@

The development of astronomical instrumentation technology in the 0.1-2 THz range has been rapidly growing in recent years. Another recent surge in interest in telecommunications THz technology has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}. For this reason, atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed. Countries near the Equator face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}. In northern South America, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}. However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the northern Andes could be suitable candidates for a world-class millimeter-wave observatory. \\

Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum \cite{southpole,radford2016,chajpwv}. The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range \cite{pardocerni,pardocerni2,cont350} due to a continuum absorption spectrum formed by collisionally broadened absorption lines of water vapor in this frequency range  \cite{linecont,submmlines,turner2009}. In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry \cite{radiometer,paine2000fourier,southpole2}, satellite measurements \cite{aqua,spacemols,suen2014,spaceradio}, radiosonde humidity measurements \cite{radiosonde,radiosonde2} or indirectly via models of in situ climatological measurements \cite{climatology} or GPS-delay studies \cite{gpsorig,gpsmap}. \\

@@

The water vapor column above a potential site can change rapidly (during the course of the day) or it can change seasonally \cite{arm2013,caumont2016}. This is why medium-to-long term monitoring of the atmosphere above a site is required before building an expensive mm/sub-mm wave radio telescope. However, even in best-case scenarios, earth-based telescopes will always be limited by atmospheric absorption of light specially at THz frequencies, where frequency windows of observations are few and in some cases, narrow \cite{arrlim,limits}. \\

@@

From the climatological clusters in Figures \ref{clusth}-\ref{clustb} we selected clusters that indicate a dry, sunny climate. Table \ref{tabclu} shows the lowest-BIC number of clusters and covariance method for each variable along with our selected clusters. From here on, if a station appears in one of our preferred clusters, we will say that it satisfies our criterion for that specific variable.\\

@@

We wish to identify all stations for whom all four criteria (humidity, precipitation, rain days, sunshine duration) for identifying a region as having a stable sunny, dry climate can be met.  However, since only a few stations measured all variables, it is not straightforward to accept or reject stations based on their satisfying a criterion, i.e. belonging to one of our preferred clusters. This requires setting up a measure of quality for individual stations depending on their location and the variables they measured. We will describe the selection process and probabilistic analysis that allows us to quantify a station's capability for indicating regions with an unusually dry, sunny climate.\\

We start by creating a shortlist of stations that meet criteria for (and only for) the variables they measured. This means that if a station has precipitation and humidity data, it should belong to a humidity and precipitation cluster listed in Table \ref{tabclu} or else it is rejected. 665 stations across Colombia showed such behavior. However, since we are interested in high-mountain regions, we included a criterion for elevation limiting our list to 119 stations located above 2000 masl. This shortlist was further reduced by requiring that the total average plus 1$\sigma$ for precipitation, rain days, and relative humidity is less than 4.3 mm/d, 17 d/mo, and 81\% respectively, and for sunshine duration, more than 4 h/d for any given station. This left us with 83 weather stations. \\

However, we want to give a measure of the quality of the station vis-a-vis our criteria, i.e. a station that meets less than all four criteria should still be considered in our analysis, but it cannot be given the same importance as a station that meets all four criteria. In order to account for this, and controlling for the fact that stations are not distributed evenly in elevation (Figure \ref{totmap}b), we propose a quality measure based on a probabilistic analysis of the data.

@@

We quantify the quality of a given site using the probability for the hypothesis that a station $j$ which appears in our shortlist meets all four criteria, given that it is located at a given elevation. We can write down this probability $P_j$ in terms of the following events:

@@

Many of our 83 candidate stations, selected using the method described at the beginning of this section, were lacking in relative humidity and sunshine duration data. Given that for stations less than 5 km away the difference in relative humidity and sunshine duration data is less than 15\%, we extrapolated using data from nearby stations (at less than 4 km away and at an elevation difference of less than 100 m). The extrapolated data evidenced that 4 of the stations in the shortlist did not satisfy the relative humidity and sunshine duration criteria.  The remaining 79 stations are mapped in Figure \ref{shortlist}a.\\

We rejected 9 of the stations in Figure \ref{shortlist}a, i.e. those located outside of the region located within the latitude, longitude range $[(4.3,6.2)^\circ,(-72.4,-74.5)^\circ]$. Even though they could be indicative of unusually dry, sunny regions, there are simply not enough stations nearby to say anything conclusively, as evidenced in Figure \ref{outliers}. Adding to this, the only variable they measured was precipitation ($Q_j=2.2$), which leads us to reject them from our shortlist. \\

@@

We selected clusters of climatologically similar stations across Colombia using an unsupervised learning low-dimensional embedding algorithm for each measured variable. From these clusters, 70 weather stations came up as potential candidates for identifying high-mountain regions (at an elevation greater that 2000 masl) with an unusually dry, sunny weather. Seasonally, the driest months of the year for all the candidate regions of interest are December-February and June-August (Figures \ref{clusth}-\ref{clustb}).

@@

Figure \ref{unif} shows 6 candidate regions of interest, where weather stations indicating dry, sunny climate are regionally correlated within a $10-20$ km radius. In order to visualize the extent of those regions we plotted the predicted variance regions for each cluster along with other (rejected) stations in our sample in Figure \ref{unif}. It should be noted that the spherical variance GMM geographical clustering overestimates the extent of each region, and tells us nothing of their actual geographical shape. \\

@@

The first cluster is located at ($4.77^\circ$N, $74.18^\circ$W) and covers an radius of 15.3 km (Figure \ref{unif}, purple circle). This is a region located to the west of Bogot\'a at an elevation above 2600 masl, with a large variety of microclimates. This explains the significant amount of stations from this region in our shortlist. However, Figure \ref{unif} shows that within this predicted variance region there are more stations that do not satisfy our criteria than stations that do, and no single sub-region can be identified. This, compounded with the fact that most (70\%) of these stations have a $Q$ value lower than 5.6 due to the lack of humidity/sunshine data, implies that at the moment this is not a strong candidate region of interest.

@@

This cluster is located at ($5.30^\circ$N, $73.80^\circ$W) and covers a radius of 11.4 km (Figure \ref{unif}, blue circle). This region, located to the NNE of Bogot\'a is a hilly region along a valley at an elevation above 2600 masl. The ratio of number of stations that do not meet our criteria to the stations in this predicted variance region is low (0.3), and almost half of these stations (46\%) have a $Q_j$ value higher than 6. This indicates a potential candidate region, although more  sunshine and humidity data are needed.

@@

This cluster is located at ($5.59^\circ$N, $73.24^\circ$W) and covers a radius of 8.29 km (Figure \ref{unif}, red circle). The GMM-predicted variance region west of Tunja is located in the historic Tunja Canton, and even though the ratio of stations not in our shortlist to stations in this predicted region is low (0.31), there are simply not enough humidity and sunshine data to make any strong conclusion about this region. However, the high elevation of some of these stations is tantalizing, which is why we will keep this as a region of interest for our next paper.

@@

This cluster is located at ($6.04^\circ$N, $72.71^\circ$W) and covers a radius of 10.52 km (Figure \ref{unif}, dark green circle). This is a \emph{p\'aramo} region, characterized by high-mountain tundra weather, with some mountains reaching an elevation of up to 3800 masl. The ratio of stations rejected stations to stations in this predicted variance region is high (0.5), and it is unlikely that the mist allows for a low-atmospheric water vapor region to be located here, even if rain is sporadic. The lack of a high-$Q$ station means that more sunshine-humidity data are needed, but this is possibly not a region of interest.

@@

We compared our results to the work of  \citeasnoun{suen2014} and \citeasnoun{suen2016}, where they obtained global precipitable water vapor maps for 2012 using data from the MODIS instrument on board the Aqua and Terra satellites. The lowest mean precipitable water vapor ($\sim15$ mm) regions in the \citeasnoun{suen2014} map for Colombia are near the Ubat\'e, Tunja, Valle del Sol, and Pisba regions. The \citeasnoun{suen2016} global median precipitable water vapor map has an upper bound of 10 mm for stations located in the clusters corresponding to those same regions. Unfortunately, the resolution of the maps is not fine enough to pinpoint a site location at a precision higher than $\sim20$ km.

@@

In this paper we find plausible locations for a high-mountain mm-wave astronomical observatory in the northern Andes of Colombia, by analizing 30 years of climate data from 2046 weather stations. By low-dimensionally embedding the data, we are able to group together and correlate climate behavior indicative of daytime dry, clear-sky weather conditions.  A repository with all our code and the full dataset is available at \texttt{https://github.com/saint-germain/ideam}.\\

@@

From those 70 stations we obtained 6 geographically correlated candidate regions of interest: La Sabana, Valle de Ubat\'e, Cant\'on de Tunja, Villa de Leyva, Valle del Sol, and Parque Nacional Natural Pisba. Seasonally, there appear to be two times of the year (Dec-Feb and Jun-Aug) where the weather conditions indicate the best daytime dry, clear-sky conditions. These months could be the best for mm-wave astronomical observations to be carried out.\\

We summarize the information about two regions of interest (Valle de Ubat\'e and Valle del Sol) in Table \ref{tabfin}, selected based on a Bayesian probabilistic quality index. Most of the stations in this table are located at elevations above 2600 masl and below 3000 masl. We validated our results using satellite measurements of the upper bound for mean precipitable water vapor \cite{suen2014,suen2016} for those two regions, which is reported to be $\sim15$ mm in $\sim20$ km resolution maps. This means that it is not implausible to find a site with an even lower water vapor column in one of the regions of interest identified here. Some mountains nearby Valle de Ubat\'e and Valle del Sol  have elevations of up to 3400 masl, and therefore warrant further measurements using radio sondes and radiometers. \\

The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low elevation (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \cite{suen2016} maps).\\

There is an additional issue with the Valle del Sol region. The high industrial activity in this region fills the air with particulate material \cite{sogphd,sogamoso}, which would nullify the suitability of this region. Besides obtaining seasonal high-resolution precipitable water vapor maps, in our next paper we will study the amount of particulate material in the reported regions of interest using MODIS (Aqua and Terra) satellite data.\\

The methods used here can be extended and adapted to similar climatological datasets in other regions. This can be a preliminary step in radio astronomy site testing in developing countries.

@@

This work was made possible by the Vicerrector\'ia de Investigaci\'on 01-2015 research grant from Universidad ECCI. We would also like to thank the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) for providing us with the dataset.

>>>>

\begin{abstract}
We set out to evaluate the potential of the Colombian Andes for millimeter-wave astronomical observations. Previous studies for astronomical site testing in this region have suggested that nighttime humidity and cloud cover conditions make most sites unsuitable for professional visible-light observations. Millimeter observations can be done during the day, but require that the precipitable water vapor column above a site stays below $\sim$10 mm. Due to a lack of direct radiometric or radiosonde measurements, we present a method for correlating climate data from weather stations to sites with a low precipitable water vapor column. We use unsupervised learning techniques to low-dimensionally embed climate data (precipitation, rain days, relative humidity, and sunshine duration) in order to group together stations with similar long-term climate behavior. The data were taken over a period of 30 years by 2046 weather stations across the Colombian territory. We find 6 regions with unusually dry, clear-sky conditions, ranging in elevations from 2200 to 3800 masl. We evaluate the suitability of each region using a quality index derived from a Bayesian probabilistic analysis of the station type and elevation distributions. Two of these regions show a high probability of having an exceptionally low precipitable water vapor column. We compared our results with global precipitable water vapor maps and find a plausible geographical correlation with regions with low water vapor columns ($\sim10$ mm) at an accuracy of $\sim20$ km. Our methods can be applied to similar datasets taken in other countries as a first step toward astronomical site evaluation.
\end{abstract}

@@

The development of astronomical instrumentation technology in the 0.1-2 THz range has been rapidly growing in recent years. Another recent surge in interest in technology at this frequency range has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}. For this reason, atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed. Countries near the Equator face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}. In northern South America, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}. However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the northern Andes could be suitable candidates for a world-class millimeter-wave observatory. \\

Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum \cite{southpole,radford2016,chajpwv}. The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range \cite{pardocerni,pardocerni2,cont350} due to a continuum absorption spectrum formed by collisionally broadened absorption lines of water vapor in this frequency range  \cite{linecont,submmlines,turner2009}. In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor (pwv) profile using remote sensing techniques such as microwave radiometry \cite{radiometer,paine2000fourier,southpole2}, satellite measurements \cite{aqua,spacemols,suen2014,spaceradio}, radiosonde humidity measurements \cite{radiosonde,radiosonde2}, or indirectly via models of in situ climatological measurements \cite{climatology} or GPS-delay studies \cite{gpsmet,gpsorig,gpsmap}. \\

@@

The water vapor column above a potential site can change rapidly (during the course of the day) or it can change seasonally \cite{arm2013,caumont2016}. This is why medium-to-long term monitoring of the atmosphere above a site is required before building an expensive mm/sub-mm wave radio telescope. However, even in best-case scenarios, earth-based telescopes will always be limited by atmospheric absorption of light specially at THz frequencies, where frequency windows of observations are few and in some cases, narrow \cite{arrlim,limits}.\\

We estimated the expected zenith opacity above the Colombian Andes (Figure \ref{opac}) at cm-mm wavelengths considering columns of 2, 10, and 20 mm of precipitable vater vapor and typical weather conditions. The elevation for the simulated sites in Figure \ref{opac} were chosen according to existing millimeter telescope sites (Pico Veleta, Plateau de Bure) \cite{mminf}. Thus, considering a best-case scenario (below 2 mm pwv), the potential observing frequency bands would be $70-115$ GHz, $120-170$ GHz, and $206-223$ GHz. The opacity models were calculated using the CASA \cite{casa} \texttt{asap.opacity.model} class, and deriving the pwv value from the 22 GHz K-band zenith opacity \cite{deuber,sanpedromartir}. \\

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{opac.pdf}
\caption{Atmospheric opacity models for the Colombian Andes for different precipitable water vapor (pwv) values corresponding to three scenarios: \textit{Red line} (20 mm pwv) a 2600 masl site with 90\% relative humidity at 13$^\circ$C, \textit{Blue line} (10 mm pwv) a 3400 masl site with 90\% relative humidity at 0$^\circ$C, \textit{Red line} (2 mm pwv) a 3400 masl site with 30\% relative humidity at 0$^\circ$C.}\label{opac}
\end{center}
\end{figure}

@@

From the climatological clusters in Figures \ref{clusth}-\ref{clustb} we selected clusters that indicate a clear-sky, potentially dry climate. Table \ref{tabclu} shows the lowest-BIC number of clusters and covariance method for each variable along with our selected clusters. From here on, if a station appears in one of our preferred clusters, we will say that it satisfies our criterion for that specific variable.\\

@@

We wish to identify all stations for whom all four criteria (low humidity, precipitation, rain days, and high sunshine duration) for identifying a region as having a clear-sky, potentially dry climate can be met.  However, since only 6\% of the stations measured all variables (type $T_8$ in Table \ref{tabstype}), it is not straightforward to accept or reject stations based on their belonging to one of our selected clusters. In other words, a station that meets less than all four criteria can still be considered in our analysis, but it will not be given the same level of importance as a station that meets all four criteria. This requires setting up a method to determine the quality of a site depending on its location and station type.  Thus we selected stations that meet criteria for (and only for) the variables they measured. Thus, if a station measured precipitation and humidity (Type $T_3$), it had to belong to a precipitation and humidity cluster listed in Table \ref{tabclu} or else it was rejected. 665 stations across Colombia showed such behavior. \\

In order to estimate how much consideration we should give to a shortlisted station with an incomplete measurement set, we propose the use of a quality index based on a probabilistic analysis of the entire dataset, accounting for the non-uniform joint distribution of station types (Table \ref{tabstype}) and elevations (Figure \ref{totmap}b). Neglecting to include all stations in the calculation of this joint probability leads to very strong selection biases, assigning unfairly high quality indices to substandard stations. Therefore, even though we will later limit our selection to high-elevation sites in order to achieve lower pwv regions, we considered all the information provided by the full dataset in our probabilistic analysis.

@@

We quantify the quality of a given site with the probability for the hypothesis that a station $j$ which appears in our shortlist meets all four criteria, given that it is located at a given elevation and is of a given type. We can write down this probability $P_j$ in terms of the following events:

@@

As mentioned above, as we go higher in elevation, the water vapor column decreases. For this reason we included an additional criterion for elevation limiting our list to 119 stations located above 2000 masl. This list was further reduced by requiring that for any given station, the total average plus 1$\sigma$ for precipitation, rain days, and relative humidity is respectively less than 4.3 mm/d, 17 d/mo, and 81\%, and for sunshine duration, more than 4 h/d. This left us with a shortlist of 83 weather stations.\\

Many of the stations in our dataset are lacking in relative humidity and sunshine duration data. In order to see if we could extrapolate using data from nearby stations, we estimated the difference between relative humidity and sunshine duration data for all stations in the entire dataset. For stations at a relative elevation of less than 100 m and located at less than 4 km away from each other, this difference amounts to less than 15\%. Thus, we extrapolated data for some of the 83 candidate stations in our shortlist, and were able to reject 4 stations which did not satisfy the relative humidity and sunshine duration criteria. The remaining 79 stations are plotted in Figure \ref{shortlist}a.\\

We rejected 9 additional stations located outside of the region located within the latitude, longitude range $[(4.3,6.2)^\circ,(-72.4,-74.5)^\circ]$ (Figure \ref{shortlist}a) due to geographical sparsity. Even though they could be indicative of clear-sky, potentially dry regions, there are not enough stations nearby to make any strong conclusions regarding those regions, as shown in Figure \ref{outliers}. Adding to this, in all 9 cases the only variable they measured was precipitation ($Q_j=2.2$), which led us to reject them from our shortlist. \\

@@

We selected clusters of climatologically similar stations across Colombia using an unsupervised learning low-dimensional embedding algorithm for each measured variable. From these clusters, 70 weather stations came up as potential candidates for identifying high-mountain regions (at an elevation greater that 2000 masl) with clear-sky, potentially dry weather. The driest months of the year for all the candidate regions of interest are December-February and June-August (Figures \ref{clusth}-\ref{clustb}). It should be noted that  humidity in the Colombian Andes can change rapidly during the course of the day \cite{pinzon}, and daytime relative humidity measurements $<50$\% are often reported in high-mountain plateau stations.

@@

Figure \ref{unif} shows 6 candidate regions of interest, where weather stations indicating clear-sky, potentially dry climate are regionally correlated within a $10-20$ km radius. In order to visualize the extent of those regions we plotted the predicted variance regions for each cluster along with other (rejected) stations in our sample in Figure \ref{unif}. It should be noted that the spherical variance GMM geographical clustering overestimates the extent of each region, and tells us nothing of their actual geographical shape. \\

@@

The first cluster is located at ($4.77^\circ$N, $74.18^\circ$W) and covers an radius of 15.3 km (Figure \ref{unif}, magenta circle). This is a region located to the west of Bogot\'a at an elevation above 2600 masl, with a large variety of microclimates. This explains the significant amount of stations from this region in our shortlist. However, Figure \ref{unif} shows that within this predicted variance region there are more stations that do not satisfy our criteria than stations that do, and no single sub-region can be identified. This, compounded with the fact that most (70\%) of these stations have a $Q$ value lower than 5.6 due to the lack of humidity/sunshine data, implies that at the moment this is not a strong candidate region of interest.

@@

This cluster is located at ($5.30^\circ$N, $73.80^\circ$W) and covers a radius of 11.4 km (Figure \ref{unif}, cyan circle). This region, located to the NNE of Bogot\'a is a hilly region along a valley at an elevation above 2600 masl. The ratio of number of stations that do not meet our criteria to the stations in this predicted variance region is low (0.3), and almost half of these stations (46\%) have a $Q_j$ value higher than 6. This indicates a potential candidate region, although more  sunshine and humidity data are needed.

@@

This cluster is located at ($5.59^\circ$N, $73.24^\circ$W) and covers a radius of 8.29 km (Figure \ref{unif}, purple circle). The GMM-predicted variance region west of Tunja is located in the historic Tunja Canton, and even though the ratio of stations not in our shortlist to stations in this predicted region is low (0.31), there are simply not enough humidity and sunshine data to make any strong conclusion about this region. However, the high elevation of some of these stations is tantalizing, which is why we will keep this as a region of interest for our next paper.

@@

This cluster is located at ($6.04^\circ$N, $72.71^\circ$W) and covers a radius of 10.52 km (Figure \ref{unif}, teal circle). This is a \emph{p\'aramo} region, characterized by high-mountain tundra weather, with some mountains reaching an elevation of up to 3800 masl. The ratio of stations rejected stations to stations in this predicted variance region is high (0.5), and it is unlikely that the mist allows for a low-atmospheric water vapor region to be located here, even if rain is sporadic. The lack of a high-$Q$ station means that more sunshine-humidity data are needed, but this is possibly not a region of interest.

@@

We compared our results to the work of  \citeasnoun{suen2014} and \citeasnoun{suen2016}, where they obtained global precipitable water vapor maps for 2012 using data from the MODIS instrument on board the Aqua and Terra satellites. Even though raw MODIS data has a potentially higher spatial resolution, the resolution of the processed maps is not fine enough to pinpoint a site location at a precision higher than $\sim20$ km. The area with the lowest mean precipitable water vapor that MODIS is able to measure ($\sim10-15$ mm) in the \citeasnoun{suen2014} map for Colombia spans the Tunja, Valle del Sol, Pisba, and part of the Valle de Ubat\'e regions. However, it also spans nearby regions with rejected stations, which means that low-resolution precipitable water vapor maps can indicate a wide, first order candidate region, but fail to account for highly local weather conditions, which are accounted for in our method. For this reason, in our next paper we will correlate the results presented here with existing GPS-delay water vapor measurements from the Colombian Global Navigation Satellite System GeoRED network (\texttt{https://geored.sgc.gov.co}) and seasonal, higher resolution MODIS data.

@@

In this paper we find plausible locations for a high-mountain cm- to mm-wave astronomical observatory in the northern Andes of Colombia, by analizing 30 years of climate data from 2046 weather stations. By low-dimensionally embedding the data, we are able to group together and correlate climate behavior indicative of daytime dry, clear-sky weather conditions.  A repository with all our code and the full dataset is available at \texttt{https://github.com/saint-germain/ideam}.\\

@@

From those 70 stations we obtained 6 geographically correlated candidate regions of interest: La Sabana, Valle de Ubat\'e, Cant\'on de Tunja, Villa de Leyva, Valle del Sol, and Parque Nacional Natural Pisba. Seasonally, there appear to be two times of the year (Dec-Feb and Jun-Aug) where the weather conditions indicate the best daytime dry, clear-sky conditions. These months could be the best for mm-wave astronomical observations to be carried out. The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low elevation (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \cite{suen2016} maps). The La Sabana region is also not a good candidate due to a high rejection rate of stations within this region.\\

We summarize the information about two regions of interest (Valle de Ubat\'e and Valle del Sol) in Table \ref{tabfin} with the highest Bayesian probabilistic indices, reflecting the quality of the data for stations in those regions. Most of the stations in this table are located at elevations above 2600 masl and below 3000 masl. We validated our results using satellite measurements of the upper bound for mean precipitable water vapor \cite{suen2014,suen2016} for those two regions, which is reported to be $\sim10-15$ mm in $\sim20$ km resolution maps. This means that it is not implausible to find a site with an even lower water vapor column in one of the regions of interest identified here. Some mountains nearby Valle de Ubat\'e and Valle del Sol have elevations of up to 3400 masl, and therefore could warrant further measurements using GPS stations, radio sondes and radiometers. The same could be said for the Pisba and Tunja regions, but the low quality of the climate data (indicated by low Bayesian probabilistic indices) indicates that more data is needed. For this reason, we will correlate the results for these four regions with GPS-delay water vapor measurements in our following paper.\\

There is an additional issue with the Valle del Sol region. The high industrial activity in this region fills the air with particulate material \cite{sogphd,sogamoso}, which would nullify the suitability of this region. Besides obtaining seasonal, higher-resolution precipitable water vapor maps, in our next paper we will study the amount of particulate material in the reported regions of interest using MODIS (Aqua and Terra) satellite data.\\

The methods used here can be extended and adapted to climatological datasets in other countries. Given a similar spatial and temporal coverage, our methods can provide a better picture of local climate than global satellite maps, and can help narrow down plausible locations for direct atmospheric opacity measurement campaigns. This can be a low-cost, preliminary step in radio astronomy site testing in developing countries.

@@

This work was made possible by the Vicerrector\'ia de Investigaci\'on 01-2015 research grant from Universidad ECCI. We would also like to thank the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) for providing us with the dataset, and the anonymous referee for helping us convey our ideas more clearly.
